---
title: "Reproducible Research Project 1"
author: "Rafael Gurgel"
date:  "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
    if(!require(knitr)){
        install.packages("knitr")
        library(knitr)
    }
    if(!require(ggplot2)){
        install.packages("ggplot2")
        library(ggplot2)
    }
    if(!require(dplyr)){
        install.packages("dplyr")
        library(dplyr)
    }
})
Sys.setlocale(category = "LC_ALL", locale = "english")   # Set to show English date format
opts_chunk$set(echo=TRUE,warning=FALSE,message=FALSE,fig.path="Figs/")
```

## Introduction
This is the Course Project 1 of the *Reproducible Research* course. It analysis a provided data from a personal activity monitoring device during the months of October and November, 2012.


### Reading and processing the data
The data is provided in a _.zip_ format. It was uncompressed getting the _activity.csv_ file included on [this current GitHub repository](https://github.com/rafagurgel/RepData_PeerAssessment1).
```{r, echo = TRUE}
data<-read.table("activity.csv", header=T, sep=",")
data<-transform(data,date=as.Date(date))
data.byday<- data %>%
    group_by(date) %>%
    summarise(Steps.Total = sum(steps)) %>%
    select(Date = date, Steps.Total)

data.bymin<- data %>%
    group_by(interval)%>%
    summarise(Steps.Mean = mean(steps, na.rm = TRUE))%>%
    select(Interval = interval, Steps.Mean)
```

### What is mean total number of steps taken per day?
```{r, echo = TRUE, fig.height = 4, fig.width = 12,fig.align='center'}
g<- ggplot(data.byday, aes(Steps.Total),na.rm=TRUE)
g + geom_histogram(color="#222222",fill="steelblue", alpha=0.8) +
    theme_bw() + 
    labs(title="Total Steps given by a single day",x ="Steps", y = "Frequency") + 
    theme(plot.title = element_text(hjust = 0.5))
# Save figure
tmp <- dev.copy(png,'figure/histogram_01.png')
tmp <- dev.off()

# Mean daily steps
steps.mean <- mean(data.byday$Steps.Total,na.rm=TRUE)
# Median daily steps
steps.median <- median(data.byday$Steps.Total,na.rm=TRUE)
```
The mean of daily steps is **`r format(steps.mean, scientific=FALSE)`** and the median is **`r format(steps.median, scientific=FALSE)`**.

### What is the average daily activity pattern?
```{r, echo = TRUE, fig.height = 4, fig.width = 12,fig.align='center'}
# Get the max point in dataframe
max.interval<- data.bymin%>%filter(Steps.Mean == max(Steps.Mean))
g<- ggplot(data.bymin, aes(Interval,Steps.Mean),na.rm=TRUE)
g + geom_line() +
    theme_bw() + 
    scale_y_continuous(breaks = c(0,50,100,150,200,max.interval$Steps.Mean), labels=c(0,50,100,150,200,sprintf("  %.2f",max.interval$Steps.Mean)   )) +
    scale_x_continuous(breaks = c(0,500,1000,1500,2000,max.interval$Interval), labels=c(0,500,1000,1500,2000,max.interval$Interval)) + 
    theme(axis.text.y = element_text(colour = c('black', 'black', 'black', 'black', 'black', 'red' )), axis.text.x = element_text(colour = c('black', 'black', 'black', 'black', 'black', 'red' ))) + 
    geom_vline(xintercept = max.interval$Interval, 'col' = 'red', linetype='dotted') + 
    geom_hline(yintercept = max.interval$Steps.Mean, 'col' = 'red',linetype='dotted') +
    geom_point(aes(max.interval$Interval, max.interval$Steps.Mean), col='red') +
    labs(title="Average steps by day Intervals",x ="Interval", y = "Mean Steps") + 
    theme(plot.title = element_text(hjust = 0.5))
# Save figure
tmp <- dev.copy(png,'figure/timeseries_01.png')
tmp <- dev.off()
```
As we can see, the user has achieved the biggest average steps of **`r format(max.interval$Steps.Mean, scientific=FALSE)`** when has passed **`r format(max.interval$Interval, scientific=FALSE)`** minutes.

### Inputing missing values

```{r,  echo = TRUE}
nna <- with(data,sum(is.na(steps)))
```
There are  **`r format(nna, scientific=FALSE)`** missing values coded as NA. These values will be replaced by the median of the 5-minute interval.

```{r, echo = TRUE, fig.height = 4, fig.width = 12,fig.align='center'}
# Get the median by interval
data.bymin.median<- data %>%
    group_by(interval)%>%
    summarise(Steps.Median = median(steps, na.rm = TRUE))%>%
    select(Interval = interval, Steps.Median)

data.byday2 <- data %>%
    mutate(Steps= ifelse(is.na(steps),data.bymin.median$Steps.Median, steps))%>%
    group_by(date) %>%
    summarise(Steps.Total = sum(Steps)) %>%
    select(Date = date, Steps.Total)

data.bymin2<- data %>%
    mutate(Steps= ifelse(is.na(steps),data.bymin.median$Steps.Median, steps))%>%
    group_by(interval)%>%
    summarise(Steps.Mean = mean(steps, na.rm = TRUE))%>%
    select(Interval = interval, Steps.Mean)
# Histogram
g<- ggplot(data.byday2, aes(Steps.Total),na.rm=TRUE)
g + geom_histogram(color="#222222",fill="steelblue", alpha=0.8) +
    theme_bw() + 
    labs(title="Total Steps given by a single day",x ="Steps", y = "Frequency") + 
    theme(plot.title = element_text(hjust = 0.5))
# Save figure
tmp <- dev.copy(png,'figure/histogram_02.png')
tmp <- dev.off()
# Mean daily steps
steps.mean2 <- mean(data.byday2$Steps.Total,na.rm=TRUE)
# Median daily steps
steps.median2 <- median(data.byday2$Steps.Total,na.rm=TRUE)
```
As we can see, the mean and median's values (**`r format(steps.mean2, scientific=FALSE)`** and **`r format(steps.median2, scientific=FALSE)`**) differ than the values calculated before (**`r format(steps.mean, scientific=FALSE)`** and **`r format(steps.median, scientific=FALSE)`**). It is expected considerating the filling strategy adopted.

### Are there differences in activity patterns between weekdays and weekends?
First we need to create a new factor variable in the dataset indicating whether a given date is a weekday or weekend day. It wasn't informed if the data in this analysis will be the original or with the NA's replacement. I decided to use the original data
```{r,  echo = TRUE}
data.bymin.weekdays <- data %>%
    group_by(date) %>%
    mutate(Weekday = as.factor(weekdays(date)))%>%
    mutate(Weekday = ifelse((Weekday == "Saturday")||(Weekday == "Sunday"),"Weekend","Weekday"))%>%
    mutate(Weekday = as.factor(Weekday))%>%
    group_by(interval,Weekday)%>%
    summarise(Steps.Mean = mean(steps, na.rm = TRUE))%>%
    select(Interval = interval, Weekday,Steps.Mean)
```

```{r, echo = TRUE, fig.height = 8, fig.width = 12,fig.align='center'}
g<- ggplot(data.bymin.weekdays, aes(Interval,Steps.Mean),na.rm=TRUE)
g + geom_line() +
    theme_bw() + 
    facet_wrap(~Weekday,ncol = 1) + 
    labs(title="Average steps by day Intervals",x ="Interval", y = "Mean Steps") + 
    theme(plot.title = element_text(hjust = 0.5),strip.text = element_text(face="bold", size=14))
# Save figure
tmp <- dev.copy(png,'figure/timeseries_02.png')
tmp <- dev.off()
```